---
title: "Week 11 · Novelty Functions and Tempograms"
author: "John Ashley Burgoyne"
date: "12 March 2025"
output: 
  html_document:
    toc: true
    theme:
      heading_font:
        google: 
          family: Rajdhani
          wght: 700
      base_font:
        google: Fira Sans
      code_font:
        google: Fira Mono
      bg: "#FFFFFF"
      fg: "#212529" 
      primary: "#2b2bee"
      secondary: "#39d7b8"
      success: "#39d7b8"
      danger: "#fa5577"
      warning: "#ffb14c"
      info: "#0cc7f1"
---

You can download the raw source code for these lecture notes [here](compmus2025-w11.Rmd).

## Course Meeting Plan

### Wednesday · 12 March · Lecture

  - Lecture: Onset detection (15 mins)
  - Breakout: Energy or spectrum? (15 mins)
  - Discussion: Breakout findings (5 mins)
  - Demo: Beat tracking (15 mins)
  - Lecture: Tempo estimation (15 mins)
  - Discussion: Preferred tempo (10 mins)
  - Portfolio critiques (15 min)
  
### Wednesday · 12 March · Novelty Functions

  - Demo: Novelty functions in Spotify (15 mins)
  - Breakout: Novelty functions (20 mins)
  - Discussion: Breakout findings (10 mins)
  - Demo: Tempograms in Spotify (15 mins)
  - Breakout: Tempograms (20 mins)
  - Discussion: Breakout findings (10 mins)

## Set-up

Again, you will want to download a new version of `compmus.R` this week.
You should also check make sure that you have the `signal` package installed, which is necessary for generating tempograms.
You do not need to load it explicitly, but the new functions in `compmus.R` will complain if it is not installed.
Use Install Packages (under the Tools menu) in RStudio to install it.

```{r, results = 'hide'}
library(tidyverse)
source("compmus.R")
```

## Breakout 1: Energy or Spectrum?

Look at one (or more) of the self-similarity matrices from somebody's portfolio in your group. Discuss what you think a spectrum-based novelty function would look like for this track. Listen to (some of) the track and also discussion what you think an energy-based novelty function would look like. Which one do you think would be most useful for beat tracking, and why? 

## Breakout 2: Novelty Functions

We can compute an energy-based novelty function based on Essentia's loudness estimates. 
Try the `compmus_energy_novelty()` function on several files in the class corpus. 
How well do they seem to work?

```{r}
"features/ahram-j-1.json" |>
  compmus_energy_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Energy Novelty")
```

We can use similar approaches for cepstrograms to get an approximation to spectral novelty, with the `compmus_spectral_novelty()` function. 
Do you find the novelty functions using this approximation to be more or less helpful than the energy novelty?

```{r}
"features/ahram-j-1.json" |>
  compmus_spectral_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Spectral Novelty")

```

## Breakout 3: Tempograms

The new `compmus_tempogram()` function generates a Fourier-based tempogram from Essentia's estimates of beat location and beat loudness, ready to plot with `geom_raster()`.
Be warned that computing tempograms can be slow!
Try increasing the window size and hop size if you find that you are waiting too long.

```{r}
"features/ahram-j-1.json" |>
  compmus_tempogram(window_size = 8, hop_size = 1, cyclic = FALSE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()
```

The textbook notes that Fourier-based tempograms tend to pick up strongly on tempo harmonics. Wrapping into a cyclic tempogram can be more informative. 

```{r}
"features/ahram-j-1.json" |>
  compmus_tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()
```

### Instructions

Return to the track you discussed in Breakout 1 (or choose a new track that somebody in your group loves). Compute regular and cyclic tempograms for this track. 
  - How well do they work? 
  - Do you see more tempo harmonics or more tempo *sub*-harmonics? Is that what you expected? Why?
  - Try other tracks as time permits, and be prepared to share your most interesting tempogram with the class.
