---
title: "Week 10 · Key and Chord Estimation"
author: "John Ashley Burgoyne"
date: "5 March 2025"
output: 
  html_document:
    toc: true
    theme:
      heading_font:
        google: 
          family: Rajdhani
          wght: 700
      base_font:
        google: Fira Sans
      code_font:
        google: Fira Mono
      bg: "#FFFFFF"
      fg: "#212529" 
      primary: "#2b2bee"
      secondary: "#39d7b8"
      success: "#39d7b8"
      danger: "#fa5577"
      warning: "#ffb14c"
      info: "#0cc7f1"
---

You can download the raw source code for these lecture notes [here](compmus2025-w10.Rmd).

## Course Meeting Plan

### Wednesday · 5 March · Lecture

  - Demo: [Chordify](https://chordify.net/chords/zager-and-evans-in-the-year-2525-eitan19611) (10 min)
  - Lecture: Chordograms (30 min)
  - Portfolio critiques (15 min)
  - Breakout: Grease and the sitar 1 (10 min)
  - Discussion: Breakout results (5 min)
  - Breakout: Grease and the sitar 2 (10 min)
  - Discussion: Breakout results (5 min)
  - Wrap-up (5 min)

### Wednesday · 5 March · Lab

  - Demo: Chordograms with the Spotify API (15 min)
  - Breakout: Chordograms (55 min)
  - Discussion: Breakout results (20 min)

## Set-up

```{r, results = 'hide'}
library(tidyverse)
source("compmus.R")
```

## Breakout 2: Chordograms

The focus of the readings this week were chord and key estimation. One set of standard templates is below: 1--0 coding for the chord templates and the Krumhansl--Kessler key profiles. 

```{r}
#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    0,    0)
minor_chord <-
  c(   1,    0,    0,    1,    0,    0,    0,    1,    0,    0,    0,    0)
seventh_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    1,    0)

major_key <-
  c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <-
  c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <-
  tribble(
    ~name, ~template,
    "Gb:7", circshift(seventh_chord, 6),
    "Gb:maj", circshift(major_chord, 6),
    "Bb:min", circshift(minor_chord, 10),
    "Db:maj", circshift(major_chord, 1),
    "F:min", circshift(minor_chord, 5),
    "Ab:7", circshift(seventh_chord, 8),
    "Ab:maj", circshift(major_chord, 8),
    "C:min", circshift(minor_chord, 0),
    "Eb:7", circshift(seventh_chord, 3),
    "Eb:maj", circshift(major_chord, 3),
    "G:min", circshift(minor_chord, 7),
    "Bb:7", circshift(seventh_chord, 10),
    "Bb:maj", circshift(major_chord, 10),
    "D:min", circshift(minor_chord, 2),
    "F:7", circshift(seventh_chord, 5),
    "F:maj", circshift(major_chord, 5),
    "A:min", circshift(minor_chord, 9),
    "C:7", circshift(seventh_chord, 0),
    "C:maj", circshift(major_chord, 0),
    "E:min", circshift(minor_chord, 4),
    "G:7", circshift(seventh_chord, 7),
    "G:maj", circshift(major_chord, 7),
    "B:min", circshift(minor_chord, 11),
    "D:7", circshift(seventh_chord, 2),
    "D:maj", circshift(major_chord, 2),
    "F#:min", circshift(minor_chord, 6),
    "A:7", circshift(seventh_chord, 9),
    "A:maj", circshift(major_chord, 9),
    "C#:min", circshift(minor_chord, 1),
    "E:7", circshift(seventh_chord, 4),
    "E:maj", circshift(major_chord, 4),
    "G#:min", circshift(minor_chord, 8),
    "B:7", circshift(seventh_chord, 11),
    "B:maj", circshift(major_chord, 11),
    "D#:min", circshift(minor_chord, 3)
  )

key_templates <-
  tribble(
    ~name, ~template,
    "Gb:maj", circshift(major_key, 6),
    "Bb:min", circshift(minor_key, 10),
    "Db:maj", circshift(major_key, 1),
    "F:min", circshift(minor_key, 5),
    "Ab:maj", circshift(major_key, 8),
    "C:min", circshift(minor_key, 0),
    "Eb:maj", circshift(major_key, 3),
    "G:min", circshift(minor_key, 7),
    "Bb:maj", circshift(major_key, 10),
    "D:min", circshift(minor_key, 2),
    "F:maj", circshift(major_key, 5),
    "A:min", circshift(minor_key, 9),
    "C:maj", circshift(major_key, 0),
    "E:min", circshift(minor_key, 4),
    "G:maj", circshift(major_key, 7),
    "B:min", circshift(minor_key, 11),
    "D:maj", circshift(major_key, 2),
    "F#:min", circshift(minor_key, 6),
    "A:maj", circshift(major_key, 9),
    "C#:min", circshift(minor_key, 1),
    "E:maj", circshift(major_key, 4),
    "G#:min", circshift(minor_key, 8),
    "B:maj", circshift(major_key, 11),
    "D#:min", circshift(minor_key, 3)
  )
```

Armed with these templates, we can make chordograms and keygrams for individual pieces. 
The new helper function `compmus_match_pitch_templates` compares the averaged chroma vectors against templates to yield a chordo- or keygram. 
(You will need to re-download `compmus.R` from Canvas and replace your old version in order for it to work properly.)
Here is an example of how to use it.

```{r}
"features/berend-b-1.json" |> 
  compmus_chroma(norm = "identity") |> 
  compmus_match_pitch_templates(
    key_templates,         # Change to chord_templates if desired
    norm = "identity",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()                                      # Change the theme?
```

### Instructions

Once you have the code running, try the following adaptations.

  1. Explore different norms and distances, as in previous weeks.
  2. Try making a chordogram instead of the keygram above.
  3. Replace the key profiles above with Temperley's proposed improvements from the reading this week. (Don't forget to re-run the chunk after you are finished.) Do the revised profiles work any better? Can you think of a way to improve the chord profiles, too?

As a reminder, here are the norm and distance combinations that tend to work well for chroma.
  
| Domain                      | Normalisation | Distance  |
| ----------------------------|---------------|-----------|
| Non-negative (e.g., chroma) | Manhattan     | Manhattan |
|                             |               | Aitchison |
|                             | Euclidean     | cosine    |
|                             |               | angular   |
|                             | Chebyshev     | [none]    |

